Agda Meets Accelerate

### 2012-11-25 paper/main.tex
cosmetics, changed a few misleading formulations

\begin{document}
%
\title{Agda Meets Accelerate}
\subtitle{Extended Abstract}
\author{Peter Thiemann\inst{1} \and Manuel M. T. Chakravarty\inst{2}}
\institute{
  University of Freiburg, Germany,\\
  \email{thiemann@informatik.uni-freiburg.de}
\and
University of New South Wales, Sydney, Australia,\\
\email{chak@cse.unsw.edu.au}
}

@@

  This raises the question whether embedded languages for data parallel programming would benefit from fully-fledged dependent types, such as those available in Agda. We explored that question by designing and implementing an Agda frontend to Accelerate, an embedded language for data parallel programming aimed at GPUs. We discuss the potential of dependent types in this domain, describe some of the limitations that we encountered, and share some insights from our preliminary implementation.
\end{abstract}
\keywords{programming with dependent types, data parallelism}
\thispagestyle{plain}
\pagestyle{plain}
%
\section{Introduction}
\label{sec:introduction}

Generative approaches to programming parallel hardware promise to
combine high-level programming models with high performance. They are
particularly attractive for targeting restricted architectures, such
as GPUs (graphics processor units), that cannot efficiently execute
code aimed at conventional multicore CPUs. Instead, GPUs require a
high degree of data parallelism, restricted control flow, and
carefully tailored data access patterns to be efficient. Previous work
---for example, Accelerator~\cite{Tarditi:2006},
Copperhead~\cite{Catanzaro:EECS-2010-124}, and
Accelerate~\cite{ChakravartyKellerLeeMcdonellGrover2011}---
demonstrates that embedded array languages with a custom code
generator can meet those GPU constraints with carefully designed language constructs.

Given a host language with an expressive type system, it is
attractive to leverage that type system to express static properties
of the embedded language. For example, Accelerate, an embedded array
language for Haskell, uses Haskell's recent support for type-level
programming like GADTs and type families in that
manner~\cite{ChakravartyKellerLeeMcdonellGrover2011}. This design
choice is important for approaches relying on runtime code generation:
compile-time faults in the embedded language should be avoided because
it corresponds to at application runtime. Moreover, static guaranties
hold the potential to improve the predictability of parallel
performance.

Dependent types are an emerging approach to certified programming, where invariants are established in the form of types and proven at compile time. Many of Haskell's type-level extensions used in Accelerate approximate various aspects of dependently-typed programming. Hence, it it is natural to ask whether fully-fledged dependent types, such as those provided by Agda, improve the specification of an embedded language like Accelerate, whether they increase the scope of static guarantees, and whether they may be leveraged to predict performance more accurately.

@@

Our investigation has the following structure. After recalling some
background on Agda and Accelerate in Section~\ref{sec:background} and
describing related work in Section~\ref{sec:related-work}, 
Section~\ref{sec:dependent-types} discusses potential uses of
dependent types in an array-oriented data parallel language like
Accelerate and how they were realized in our
implementation. Section~\ref{sec:limitations} considers conceptual 
problems and limitations that we ran into when constructing the Agda
frontend for Accelerate. Section~\ref{sec:implementation} explains
some technical details of the implementation and discusses some
example code.

@@

The access operation takes a vector of length \verb+n+, an index
\verb+m+, and a proof of \verb+m < n+ (a derivation tree) to produce
an element of the vector.  
\begin{verbatim}
get : {A : Set} {n : Nat} -> Vec A n -> (m : Nat) -> m < n -> A
get []        _       ()      -- impossible case
get (x :: xs) zero    p       = x
get (x :: xs) (suc m) (s<s p) = get xs m p
\end{verbatim}
This code cannot fail at run time because a caller has to
construct the proof tree for \verb+m < n+ before invoking
\verb+get+. (In Agda, arguments in curly braces are \emph{implicit
  arguments} that will be inferred if omitted in an application.)
Thus, an ``index out of bounds'' error cannot happen.

@@

As computation representations of type \verb+Acc a+ are compiled at application runtime, all \verb+Acc+ compilation errors are effectivly \emph{runtime errors} of the application. Hence, Accelerate uses a range of Haskell type system extensions to statically type Accelerate expressions, such that these runtime errors are avoided where possible. In particular, Accelerate uses GADTs\cite{PeytonJonesVytiniotisWeirichWashburn2006}, associated types
\cite{ChakravartyKellerJones2005}, and type families
\cite{SchrijversPeytonJonesChakravartySulzmann2008}.

As a simple example of an Accelerate program, consider a function implementing a dot product:
%
\begin{hcode}
dotp :: Vector Float -> Vector Float -> Acc (Scalar Float)
dotp xs ys = let xs' = use xs
                 ys' = use ys
             in  fold (+) 0 (zipWith (*) xs' ys')
\end{hcode}
%
The types \verb+Vector+ and \verb+Scalar+ represent one- and zero-dimensional
arrays. Plain arrays, such as \verb+Vector Float+ are conventional Haskell arrays, using an unboxed representation to improve performance. However, when they are wrapped into the constructor \verb+Acc+, such as in \verb+Acc (Scalar Float)+, they represent arrays of the embedded language and are allocated in GPU memory, which in current high-performance GPUs is physically separate from CPU memory.

The \verb+use+ operation makes a Haskell array available in the embedded language by wrapping it into the \verb+Acc+ constructor and copying it to GPU memory.\footnote{Accelerate employs a caching strategy to avoid the transfer of arrays, which are already available in GPU memory.} The operations \texttt{fold} and \texttt{zipWith} represent collective operations on Accelerate arrays, effectively producing a representation of an array computation yielding a value of
\texttt{Scalar Float}; i.e., a single float value. This code relies
heavily on (type class) overloading: \texttt{0}, \texttt{(+)}, and
\texttt{(*)} are overloaded to just construct abstract syntax.

@@

The general type for \verb+use+ is
%
\begin{hcode}
use :: Elt e => Array sh e -> Acc (Array sh e)
\end{hcode}
%
where the class \verb+Elt+ characterises all types that may be held in Accelerate arrays. These are currently primitive types and tuples.

Common dimensions, such as \verb+DIM0+, \verb+DIM1+, and so on, are
predefined, but to enable shape polymorphic computations, along the
lines pioneered in the Haskell array library
Repa~\cite{keller-etal:repa}, shapes are inductively defined using
type-level snoc lists built from the data types \texttt{Z} and \texttt{:.}:
%
\begin{hcode}
data Z       = Z
data sh :. i = sh :. i
\end{hcode}
%
Hence, the definitions of the dimensions:
%
\begin{hcode}
type DIM0 = Z
type DIM1 = DIM0 :. Int
-- <and so on>
\end{hcode}

@@

Peebles formalised parts of the API of the Haskell array library Repa using Agda~\cite{peebles:derpa}. The formalisation uses the same shape structure as we are using in Accelerate, but array computations are neither embedded nor can parallel high-performance code be generated.

@@

Dependent ML is an ML dialect with a restricted form of dependent types, which, among other applications, may be used to statically check array bounds in array operations~\cite{xi:dml-jfp}. However, only simple indexing and array updating are considered and not aggregate array operations, such as those provided by Accelerate.

Accelerator~\cite{Tarditi:2006} enables embedded GPU computations in C\# programs; it subsequently also added F\# support. However, no attempt is made to track properties of array programs statically. Similarly, Copperhead~\cite{Catanzaro:EECS-2010-124} embedded an array language into Python, but does not attempt to track information statically.

@@

Accelerate supports a wide range of numeric types, characterized by
type class \texttt{Elt}, as base types for
array computations. Almost all of these types have no suitable
counterpart in Agda, which only supports natural numbers in unary
encoding. For that reason, our embedding keeps the Haskell types
abstract in Agda. To specify type signatures and in particular
functions that are polymorphic in such a Haskell type or depend on it
in some way, we have reified these types in an Agda type
\texttt{Element}.
\begin{verbatim}
data Element : Set where
  Bool   : Element
  Int    : Element
  Float  : Element
  Double : Element
  Pair   : Element -> Element -> Element
  -- and so on
\end{verbatim}
Corresponding to Haskell type classes that are used in Accelerate, our
embedding supplies predicates that characterize subsets. For example,
the set of numeric types is defined by a predicate
\texttt{IsNumeric}.\footnote{$\top$ is a one-element type,
  whereas $\bot$ is a type without elements. These types customarily
  represent truth and falsity.}
\begin{alltt}
IsNumeric : Element -> Set
IsNumeric Int = \(\top\)
IsNumeric Float = \(\top\)
IsNumeric Double = \(\top\)
IsNumeric _ = \(\bot\)
\end{alltt}
The embedding declares further subsets all in the same style.

@@

To see the Agda embedding in action, we translate the
dot product example from Section~\ref{sec:accelerate} to Agda.
\begin{code}
dotp : forall {E : Element} {{p : IsNumeric E}} {n : Nat}
     -> PreVector n E -> PreVector n E -> Scalar E
dotp{E} xs ys = 
  let xs' = use xs
      ys' = use ys
  in
  fold _+_ ("0" ::: E) (zipWith _*_ xs' ys')
\end{code}
Unlike the Accelerate code, this function is polymorphic with respect
to the array element type, provided it is numeric.\footnote{In Agda, arguments in double curly braces are \emph{instance arguments}~\cite{DevriesePiessens2011}. We use them much like type class constraints are used in Haskell.} It furthermore
takes the length $n$ as a parameter thus ensuring that the two input
vectors have the same size. The \texttt{PreVector} type of the
arguments corresponds to the plain \texttt{Vector} type in Accelerate,
whereas the result type \texttt{Scalar E} corresponds to \texttt{Acc
  (Scalar E)}---a piece of abstract syntax.

The \texttt{use} function works as before, but its type
includes more information:
\begin{code}
use : {sh : Shape}{E : Element} -> PreArray sh E -> Array sh E
\end{code}
Like \texttt{E}, the index \texttt{sh} is now an element of an
ordinary type instead of having to rely on type-level snoc
lists:\footnote{Recent work on Haskell's type system manages to avoid
  this issue \cite{YorgeyWeirichCretinJonesVytiniotisMagalhaes2012}.}
\begin{verbatim}
data Shape : Set where
  Z     : Shape
  _:<_> : Shape -> Nat -> Shape
\end{verbatim}
Asking for arrays of equal shape, as in the signature of \texttt{use},
means that the arrays have to have the exact same layout. The
\texttt{PreVector} and \texttt{Vector} types are just synonyms as in
Haskell:
\begin{verbatim}
PreVector n E = PreArray (Z :< n >) E
Vector n E    = Array    (Z :< n >) E
\end{verbatim}
The functions \texttt{fold}, \texttt{zipWith}, and \texttt{:::} are
discussed in the subsequent subsections. The functions \verb|_+_| and
\verb|_*_| both have the same type:
\begin{code}
_+_ : {E : Element} {{p : IsNumeric E}} -> Exp E -> Exp E -> Exp E
\end{code}
They are restricted to arguments of numeric type and construct
abstract syntax for an addition or a multiplication by delegating to
the corresponding Accelerate functions. The type \texttt{Exp E},
which denotes an AST of an expression of type \texttt{E}.

@@

Accelerate's API features expressive type constraints that describe
the shape of the array arguments and results. These constraints ensure
that no shape mismatches occur (e.g., a one-dimensional array cannot be
considered two-dimensional). However, they do not ensure at compile
time that the sizes of the dimensions match up.

@@

As \texttt{Shape} is an ordinary data type in Agda, we can define a
\texttt{size} function that computes the number of elements stored in
an array of a certain shape.
\begin{verbatim}
size : Shape -> Nat
size Z = 1
size (sh :< n >) = size sh * n
\end{verbatim}
Now we can state an accurate type for \texttt{reshape} in Agda, which
involves an extra argument with a proof that the source and
target shapes have the same size.
\begin{alltt}
reshape : \{sh : Shape\} \{E : Element\}
       -> (sh' : Shape) -> Array sh E -> (size sh \(\equiv\) size sh')
       -> Array sh' E
\end{alltt}
There is a subtle difference to the original signature. In
Accelerate, the first argument is an \emph{expression} that produces a
value of type \texttt{sh} at run time, whereas the Agda
\texttt{reshape} requires a \texttt{Shape} as its first argument.

Furthermore, functions like \texttt{map} and \texttt{zipWith} obtain
more precise types. The type of \texttt{map} tells us that the input
shape is identical to the output shape:
\begin{verbatim}
map : {A B} {sh} -> (Exp A -> Exp B) -> Array sh A -> Array sh B
\end{verbatim}
Similarly, the type of \texttt{zipWith} restricts its input arrays to
identical shapes:
\begin{verbatim}
zipWith : {A B C} {sh} -> (Exp A -> Exp B -> Exp C)
        -> Array sh A -> Array sh B -> Array sh C
\end{verbatim}
The latter type is more restrictive than the Accelerate
implementation of \texttt{zipWith}. Instead of checking the sizes
of the input arrays, it truncates them to the respective minima. A
corresponding Agda type could be developed easily. It just requires a
binary function that computes the minimum of two shapes, which is a
simple exercise.

@@

Some parallel reduction operations require their base operation to be
associative to return a predictable result. Here are two examples from
Accelerate. 
\begin{hcode}
fold  :: (Shape ix, Elt a) =>
         (Exp a -> Exp a -> Exp a) -> Exp a ->
         Acc (Array (ix :. Int) a) -> Acc (Array ix a)
fold1 :: (Shape ix, Elt a) =>
         (Exp a -> Exp a -> Exp a) ->
         Acc (Array (ix :. Int) a) -> Acc (Array ix a)
\end{hcode}
In both cases, the text of the documentation says that ``the first
argument needs to be associative'' and the \texttt{fold1}
documentation ``requires the reduced array to be non-empty''.
The second requirement can be enforced by asking for a suitable proof
object on each call of \texttt{fold1}:
\begin{verbatim}
fold1 : ... -> Array (sh :< n >) E -> (size sh * n > 0)
            -> Array sh E
\end{verbatim}
The first requirement can be rephrased to saying that the first two
parameters of \texttt{fold} together form a monoid, which requires an
associative operation with a unit element. The concept of a monoid
can be formalized in Agda, which has indeed been done in the standard
library. Unfortunately, the formalization from the library cannot be
used because Accelerate deals with ASTs, not with values. So, a
formalization is required that states that the meaning of an
AST-encoded function is associative and the meaning of another
AST-encoded constant is its unit element. Given that Accelerate
encodes AST construction using higher-order abstract syntax, such a
formalization is not straightforward. Moreover, even given expressions
with a fixed meaning, there is no general shape for associative
functions, so that proofs can only be done for special cases.

In any case, providing such information would be done by including an
additional argument that holds a suitable proof object, as in
\begin{verbatim}
fold : forall {E}{sh}{n} -> (f : Exp E -> Exp E -> Exp E) -> (e : Exp E)
     -> Array (sh :< n >) E -> IsMonoid f e -> Array sh E
\end{verbatim}
where
\begin{verbatim}
IsMonoid : forall {E} -> (f : Exp E -> Exp E -> Exp E) -> (e : Exp E) -> Set
IsMonoid f e = ( IsAssociative f , IsUnit f e)
\end{verbatim}

@@

Because of Agda's limited support for numeric data types,
we embed more ambitious numeric literals for floating point numbers
using a string with an explicit type annotation that determines the
parsing of the string. Here are some example embeddings:
\begin{verbatim}
"3.1415926" ::: Float
"6.0221415E23" ::: Double
\end{verbatim}
Recall that \texttt{Float} and \texttt{Double} are not
types, but rather values of type \texttt{Element}.
All magic of the embedding is hidden in the \texttt{:::} operation:
\begin{verbatim}
_:::_ : (s : String) -> (E : Element) 
     -> {{nu : IsNumeric E}} -> {p : T (s parsesAs E)} -> Exp E
s ::: E = Ex (constantFromString (EltDict E) (ReadDict E) s)
\end{verbatim}
The arguments \texttt{s} and \texttt{E} are explicit, but the
remaining ones are to be inferred by Agda.
As mentioned before, the argument \texttt{nu} is an instance argument; it is automatically
filled-in with a suitably typed value that is currently in scope
\cite{DevriesePiessens2011}. Here, the predicate 
\texttt{IsNumeric} plays the role of a type class that characterizes
the numeric types.

The function \texttt{parsesAs} dispatches on its ``type'' argument and
checks whether the string is a constant of the expected type. The
function \texttt{constantFromString} is imported from Accelerate.
It is an overloaded function that requires two type dictionaries,
which are computed from \texttt{E} using the functions \texttt{EltDict}
and \texttt{ReadDict}.

@@

In a number of places, Accelerate's generativity limits the
applicability of dependent typing. We already mentioned that the
formalization of associativity or of the concept of a monoid gets
unmanageable because such properties have to be asserted for abstract syntax.

For a related problem, consider an implementation of the \texttt{filter}
operation that takes a predicate and a source array and returns an
array that only contains the elements of the source array fulfilling
the predicate.  First of all, filtering only makes sense for
one-dimensional arrays, that is, for vectors. To see the second catch,
let's try to write down a dependent type signature for \texttt{filter}.
\begin{verbatim}
filter : forall {n m : Nat}{E : Element}
       -> Vector n E -> (Exp E -> Exp Bool) -> Vector m E
\end{verbatim}
The problem is that the size of the result
cannot be determined statically. In fact, the only thing we know about
\texttt{m} is that it must be less than or equal to
\texttt{n}. However, we cannot prove this from the code because of a staging restriction. The Accelerate implementation computes the length of the result only when the generated GPU code is executed. The function \texttt{Exp E -> Exp Bool} maps abstract syntax to abstract syntax; it does not directly implement a Boolean predicate. Hence, we cannot use the predicate in the result type of \texttt{filter} to more accurately constrain the size of the resulting vector --- unless we include an evaluator for Accelerate expressions.

We might contemplate emplying an existential type like
\begin{verbatim}
exists Nat (\ m -> m <= n -> Vector m E)
\end{verbatim}
However, it is not possible to build such an
existential package because the evidence \texttt{m} is not available
when the package would have to be constructed.

>>>>

\begin{document}
%
\title{Agda Meets Accelerate}
% \subtitle{Extended Abstract}
\author{Peter Thiemann\inst{1} \and Manuel M. T. Chakravarty\inst{2}}
\institute{
  University of Freiburg, Germany,\\
  \email{thiemann@informatik.uni-freiburg.de}
\and
University of New South Wales, Sydney, Australia,\\
\email{chak@cse.unsw.edu.au}
}

@@

  This observation raises the question whether an embedded language
  for data parallel programming would benefit from fully-fledged
  dependent types, such as those available in Agda. We explored that
  question by designing and implementing an Agda frontend to
  Accelerate, a Haskell-embedded language for data parallel programming aimed at GPUs. We discuss the potential of dependent types in this domain, describe some of the limitations that we encountered, and share some insights from our preliminary implementation.
\end{abstract}
\keywords{programming with dependent types, data parallelism}
\thispagestyle{plain}
\pagestyle{plain}
%
\section{Introduction}
\label{sec:introduction}

Generative approaches to programming parallel hardware promise to
combine high-level programming models with high performance. They are
particularly attractive for targeting restricted architectures that cannot efficiently execute
code aimed at conventional multicore CPUs. One prime example are GPUs
(graphics processor units), which require a
high degree of data parallelism, restricted control flow, and
custom tailored data access patterns to be efficient. Previous work
---for example, Accelerator~\cite{Tarditi:2006},
Copperhead~\cite{Catanzaro:EECS-2010-124}, and
Accelerate~\cite{ChakravartyKellerLeeMcdonellGrover2011}---
demonstrates that embedded array languages with a custom code
generator can meet those GPU constraints with carefully designed language constructs.

Given a host language with an expressive type system, it is
attractive to leverage that type system to express static properties
of the embedded language. For example, Accelerate, an embedded array
language for Haskell, uses Haskell's recent support for type-level
programming like GADTs and type families in that
manner~\cite{ChakravartyKellerLeeMcdonellGrover2011}. This design
choice is desirable for approaches relying on run-time code
generation:
each  potential fault at application run time should be discovered by
a compile-time fault in the embedded language. Moreover, static guarantees
hold the potential to improve the predictability of parallel
performance.

Dependent types are an emerging approach to certified programming, where invariants are established in the form of types and proven at compile time. Many of Haskell's type-level extensions used in Accelerate approximate aspects of dependently-typed programming. Hence, it is natural to ask whether fully-fledged dependent types, such as those provided by Agda, improve the specification of an embedded language like Accelerate, whether they increase the scope of static guarantees, and whether they may be leveraged to predict performance more accurately.

@@

Our investigation has the following structure. After recalling some
background on Agda and Accelerate in Section~\ref{sec:background} and
describing related work in Section~\ref{sec:related-work}, 
Section~\ref{sec:dependent-types} discusses potential uses of
dependent types in an array-oriented data parallel language and how they were realized in our
implementation. Section~\ref{sec:limitations} considers conceptual 
problems and limitations that we ran into when constructing the Agda
frontend for Accelerate. Section~\ref{sec:implementation} explains
some technical details of the implementation and discusses some
example code.

@@

The access operation takes a vector of length \verb+n+, an index
\verb+m+, and a proof of \verb+m < n+ (a derivation tree) to produce
an element of the vector.  
\begin{verbatim}
get : {A : Set} {n : Nat} -> Vec A n -> (m : Nat) -> m < n -> A
get []        _       ()      -- impossible case
get (x :: xs) zero    z<s     = x
get (x :: xs) (suc m) (s<s p) = get xs m p
\end{verbatim}
This code cannot fail at run time because a caller has to
construct the proof tree for \verb+m < n+ before invoking
\verb+get+. (In Agda, arguments in curly braces are \emph{implicit
  arguments} that will be inferred if omitted in an application.)
Thus, an ``index out of bounds'' error cannot happen.

@@

As computation representations of type \verb+Acc a+ are compiled at
application run time, all \verb+Acc+ compilation errors are effectively \emph{run-time errors} of the application. Hence, Accelerate uses a range of Haskell type system extensions to statically type Accelerate expressions, such that these run-time errors are avoided where possible. In particular, Accelerate uses GADTs\cite{PeytonJonesVytiniotisWeirichWashburn2006}, associated types
\cite{ChakravartyKellerJones2005}, and type families
\cite{SchrijversPeytonJonesChakravartySulzmann2008}.

As a simple example of an Accelerate program, consider a function implementing a dot product:
%
\begin{hcode}
dotp :: Vector Float -> Vector Float -> Acc (Scalar Float)
dotp xs ys = let { xs' = use xs; ys' = use ys }
             in  fold (+) 0 (zipWith (*) xs' ys')
\end{hcode}
%
The types \verb+Vector+ and \verb+Scalar+ represent one- and zero-dimensional
arrays. Plain arrays, such as \verb+Vector Float+ are conventional Haskell arrays, using an unboxed representation to improve performance. However, when they are wrapped into the constructor \verb+Acc+, such as in \verb+Acc (Scalar Float)+, they represent arrays of the embedded language and are allocated in GPU memory, which in current high-performance GPUs is physically separate from CPU memory.

The \verb+use+ operation makes a Haskell array available in the
embedded language by wrapping it into the \verb+Acc+ constructor. It
amounts to copying it to GPU memory.\footnote{Accelerate employs a
  caching strategy to avoid the transfer of arrays, which are already
  available in GPU memory.} The operations \texttt{fold} and
\texttt{zipWith} represent collective operations on Accelerate arrays,
effectively producing a representation of an array computation
yielding  a single float value (\texttt{Scalar Float}). The code relies on (type class) overloading: \texttt{0}, \texttt{(+)}, and
\texttt{(*)} are overloaded to construct abstract syntax.

@@

The general type for \verb+use+ is
%
\begin{hcode}
use :: Elt e => Array sh e -> Acc (Array sh e)
\end{hcode}
%
where the class \verb+Elt+ characterizes all types that may be held in Accelerate arrays. These are currently primitive types and tuples.

Common dimensions, such as \verb+DIM0+, \verb+DIM1+, and so on, are
predefined, but to enable shape polymorphic computations, along the
lines pioneered in the Haskell array library
Repa~\cite{keller-etal:repa}, shapes are inductively defined using
type-level snoc lists built from the data types \texttt{Z} and \texttt{:.}:
%
\begin{hcode}
data Z       = Z
data sh :. i = sh :. i
\end{hcode}
%
Hence, the definitions of the dimensions:
%
\begin{hcode}
type DIM0 = Z
type DIM1 = DIM0 :. Int
-- and so on
\end{hcode}

@@

Peebles formalized parts of the API of the Haskell array library Repa
using Agda~\cite{peebles:derpa}. The formalisation relies on the same
shape structure as Accelerate, but array computations are neither
embedded nor can parallel high-performance code be generated.

@@

Dependent ML is an ML dialect with a restricted form of dependent types, which, among other applications, may be used to statically check array bounds~\cite{xi:dml-jfp}. However, only simple indexing and array updating are considered and not aggregate array operations, such as those provided by Accelerate.

Accelerator~\cite{Tarditi:2006} enables embedded GPU computations in C\# programs; it subsequently also added F\# support. However, no attempt is made to track properties of array programs statically. Similarly, Copperhead~\cite{Catanzaro:EECS-2010-124} embeds an array language into Python, but does not attempt to track information statically.

@@

Accelerate supports a wide range of numeric types, characterized by
type class \texttt{Elt}, as base types for
array computations. Almost all of these types have no suitable
counterpart in Agda, which only supplies  computationally expensive
encodings for natural and rational numbers. For that reason, our embedding keeps the Haskell types 
abstract in Agda. To specify the types of functions that are polymorphic in such a Haskell type or depend on it
in some way, we have reified the possible element types in an Agda type
\texttt{Elt}.
\begin{verbatim}
data Elt : Set where
  Bool   : Elt
  Int    : Elt
  Float  : Elt
  Double : Elt
  Pair   : Elt -> Elt -> Elt
  -- and so on
\end{verbatim}
Corresponding to Haskell type classes that are used in Accelerate, our
embedding supplies predicates that characterize subsets. For example,
the set of numeric types is defined by a predicate
\texttt{Numeric}.\footnote{$\top$ is a one-element type,
  whereas $\bot$ is a type without elements. These types customarily
  represent truth and falsity.}
\begin{alltt}
Numeric : Elt -> Set
Numeric Int = \(\top\)
Numeric Float = \(\top\)
Numeric Double = \(\top\)
Numeric _ = \(\bot\)
\end{alltt}
The embedding declares further subsets all in the same style.

@@

To demonstrate the Agda embedding in action, we translate the
dot product example from Section~\ref{sec:accelerate} to Agda.
\begin{code}
dotp : forall {E : Elt} {{p : Numeric E}} {n : Nat}
     -> PreVector n E -> PreVector n E -> Scalar E
dotp{E} xs ys = 
  let xs' = use xs
      ys' = use ys
  in  fold _+_ ("0" ::: E) (zipWith _*_ xs' ys')
\end{code}
Unlike the Accelerate code, this function is polymorphic with respect
to the array element type, provided it is numeric.\footnote{In Agda, arguments in double curly braces are \emph{instance arguments}~\cite{DevriesePiessens2011}. We use them much like type class constraints are used in Haskell.} It furthermore
takes the length $n$ as an implicit parameter thus ensuring that the two input
vectors have the same size. The \texttt{PreVector} type of the
arguments corresponds to the plain \texttt{Vector} type in Accelerate,
whereas the result type \texttt{Scalar E} corresponds to \texttt{Acc
  (Scalar E)}---a piece of abstract syntax.

The \texttt{use} function works as before, but its type
includes more information:
\begin{code}
use : {sh : Shape}{E : Elt} -> PreArray sh E -> Array sh E
\end{code}
Like \texttt{E}, the index \texttt{sh} is now an element of an
ordinary type instead of having to rely on type-level snoc
lists:\footnote{Recent work on Haskell's type system manages to avoid
  this issue \cite{YorgeyWeirichCretinJonesVytiniotisMagalhaes2012}.}
\begin{verbatim}
data Shape : Set where
  Z     : Shape
  _:<_> : Shape -> Nat -> Shape
\end{verbatim}
Asking for arrays of equal shape, as in the signature of \texttt{use},
means that the arrays have to have the exact same layout. The
\texttt{PreVector} and \texttt{Vector} types are just synonyms as in
Haskell:
\begin{verbatim}
PreVector n E = PreArray (Z :< n >) E
Vector n E    = Array    (Z :< n >) E
\end{verbatim}
The functions \texttt{fold}, \texttt{zipWith}, and \texttt{:::} are
discussed in the subsequent subsections. The functions \verb|_+_| and
\verb|_*_| both have the same type:
\begin{code}
_+_ _*_ : {E : Elt} {{p : Numeric E}} -> Exp E -> Exp E -> Exp E
\end{code}
They are restricted to arguments of numeric type and construct
abstract syntax for an addition or a multiplication by delegating to
the corresponding Accelerate functions. The type \texttt{Exp E}
denotes an AST of an expression of type \texttt{E}.

@@

Accelerate's API features expressive type constraints that describe
the shape of the array arguments and results. These constraints ensure
that no shape mismatches occur (e.g., a one-dimensional array cannot be
considered two-dimensional), but they do not ensure at compile
time that the sizes of the dimensions match up. Such a mismatch
results in a run-time error.

@@

As \texttt{Shape} is an ordinary data type in Agda, we can define a
\texttt{size} function that computes the number of elements stored in
an array of a certain shape.
\begin{verbatim}
size : Shape -> Nat
size Z = 1
size (sh :< n >) = size sh * n
\end{verbatim}
Now we can state an accurate type for \texttt{reshape} in Agda, which
involves an extra argument with a proof that the source and
target shapes have the same size.
\begin{alltt}
reshape : \{sh : Shape\} \{E : Elt\}
       -> (sh' : Shape) -> Array sh E -> (size sh \(\equiv\) size sh')
       -> Array sh' E
\end{alltt}
There is a subtle difference to the original signature. In
Accelerate, the first argument is an \emph{expression} that produces a
value of type \texttt{sh} at run time, whereas the Agda
\texttt{reshape} requires a \texttt{Shape} as its first argument.

Furthermore, functions like \texttt{map} and \texttt{zipWith} obtain
more precise types. The type of \texttt{map} tells us that the input
shape is identical to the output shape:
\begin{verbatim}
map : {A B} {sh} -> (Exp A -> Exp B) -> Array sh A -> Array sh B
\end{verbatim}
Similarly, the type of \texttt{zipWith} restricts its input arrays to
identical shapes:
\begin{verbatim}
zipWith : {A B C} {sh} -> (Exp A -> Exp B -> Exp C)
        -> Array sh A -> Array sh B -> Array sh C
\end{verbatim}
The latter type is more restrictive than the Accelerate implementation
of \texttt{zipWith}. Instead of checking the sizes of the input
arrays, it truncates them to the respective minima. We also developed
an Agda type that directly corresponds to this implementation. It
requires a binary function that computes the minimum of two shapes,
which we leave as an exercise to the reader.

@@

Some parallel reduction operations require their base operation to be
associative to return a predictable result. Here are two examples from
Accelerate. 
\begin{hcode}
fold  :: (Shape ix, Elt a) =>
         (Exp a -> Exp a -> Exp a) -> Exp a ->
         Acc (Array (ix :. Int) a) -> Acc (Array ix a)
fold1 :: (Shape ix, Elt a) =>
         (Exp a -> Exp a -> Exp a) ->
         Acc (Array (ix :. Int) a) -> Acc (Array ix a)
\end{hcode}
In both cases, the text of the documentation says that ``the first
argument needs to be associative'' and furthermore the \texttt{fold1}
documentation ``requires the reduced array to be non-empty''.
The second requirement can be enforced by asking for a suitable proof
object on each call of \texttt{fold1}:
\begin{verbatim}
fold1 : ... -> Array (sh :< n >) E -> (size sh * n > 0)
            -> Array sh E
\end{verbatim}
The first requirement can be rephrased to saying that the first two
parameters of \texttt{fold} together form a monoid, which requires an
associative operation with a unit element. The concept of a monoid
can be formalized in Agda, which has indeed been done in the standard
library. Unfortunately, the formalization from the library cannot be
used because Accelerate deals with ASTs, not with values. So, a
formalization is required that states that the meaning of an
AST-encoded function is associative and the meaning of another
AST-encoded constant is its unit element. Given that Accelerate
encodes AST construction using higher-order abstract syntax, such a
formalization is not straightforward. Moreover, even given expressions
with a fixed meaning, there is no general shape for associative
functions, so that proofs can only be done for special cases.

In any case, providing such information would be done by including an
additional argument that holds a suitable proof object, as in
\begin{verbatim}
fold : forall {E}{sh}{n}
     -> (f : Exp E -> Exp E -> Exp E) -> (e : Exp E)
     -> Array (sh :< n >) E -> IsMonoid f e -> Array sh E
\end{verbatim}
where
\begin{verbatim}
IsMonoid : forall {E} -> (Exp E -> Exp E -> Exp E) -> Exp E -> Set
IsMonoid f e = ( IsAssociative f , IsUnit f e)
\end{verbatim}

Some readers may object that neither addition nor multiplication of
floating point numbers is associative
\cite{DBLP:journals/csur/Goldberg91}. However, for advanced
optimizations, the exploitation of algebraic laws is a necessity and
the involved degradation of precision or change of result is accepted
or account for in the error estimates. Moreover, there are
other operations, like min or max, that are commonly used with
\texttt{fold}-like operations, which are truly associative. Last, but
not least, the associativity declarations serve as an important
documentation that passing an inherently non-associative function will
produce unpredictable, implementation-dependent results.

@@

Because of Agda's limited support for numeric data types,
we embed numeric literals for integers and floating point numbers
using a string with an explicit type annotation that determines the
parsing of the string. Here are some example embeddings:
\begin{verbatim}
"3.1415926" ::: Float
"6.0221415E23" ::: Double
\end{verbatim}
Recall that \texttt{Float} and \texttt{Double} are not
types, but rather values of type \texttt{Elt}.
The \texttt{:::} operation is the workhorse of the embedding:
\begin{verbatim}
_:::_ : (s : String) -> (E : Elt) 
     -> {{nu : Numeric E}} -> {p : T (s parsesAs E)} -> Exp E
s ::: E = Ex (constantFromString (EltDict E) (ReadDict E) s)
\end{verbatim}
The arguments \texttt{s} and \texttt{E} are explicit, but the
remaining ones are inferred by Agda.
As mentioned before, the argument \texttt{nu} is an instance argument; it is automatically
filled-in with a suitably typed value that is currently in scope
\cite{DevriesePiessens2011}. Here, the predicate 
\texttt{Numeric} plays the role of a type class that characterizes
the numeric types.

The function \texttt{parsesAs} dispatches on its ``type'' argument and
parses the string to check whether it is an acceptable literal of the expected type. The
function \texttt{constantFromString} is imported from Accelerate.
It is an overloaded function that requires two type dictionaries,
which are computed from \texttt{E} using the functions \texttt{EltDict}
and \texttt{ReadDict}.

@@

In a number of places, Accelerate's generativity limits the
applicability of dependent typing. We already mentioned that the
formalization of associativity or of the concept of a monoid cannot be
verified in Agda because such properties have to be asserted for abstract syntax.

For a related problem, consider an implementation of the \texttt{filter}
operation that takes a predicate and a source array and returns an
array that only contains the elements of the source array fulfilling
the predicate.  First of all, filtering only makes sense for
one-dimensional arrays, that is, for vectors. To see the second catch,
let's try to write down a dependent type signature for \texttt{filter}.
\begin{verbatim}
filter : forall {n m : Nat}{E : Elt}
       -> Vector n E -> (Exp E -> Exp Bool) -> Vector m E
\end{verbatim}
The problem is that the size of the result
cannot be determined statically. In fact, the only thing that we could
prove from an implementation of \texttt{filter} is that
\texttt{m} must be less than or equal to \texttt{n}.
However, we cannot even prove this restriction on \texttt{m} for two
reasons. First, the code of \texttt{filter} is not available as even
the underlying Haskell library only deals with abstract
syntax. Second, even if the code were available, there is still a
problem due to staging. The Accelerate implementation computes the length of the result only when the generated GPU code is executed. The function \texttt{Exp E -> Exp Bool} maps abstract syntax to abstract syntax; it does not directly implement a Boolean predicate. Hence, we cannot use the predicate in the result type of \texttt{filter} to more accurately constrain the size of the resulting vector --- unless we include an evaluator for Accelerate expressions.
